Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires

### 2025-09-02 NorthDakotaBurnSeverity/writeup/SeverityFireBehavior.Rmd
ND severity paper first draft

---
title: 'Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires'
subtitle:  'Figures and results'
author: "DAM"
date: "Last compiled: `r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning= FALSE,
                      message= FALSE, eval = TRUE, 
                      fig.path = "./", dev = c('png', 'pdf')) 
pacman::p_load(tidyverse, sf, ggpattern, pander, readxl, grid, gridExtra)
pacman::p_load_gh('devanmcg/wesanderson')
```

@@

@@

reg_gg <- 
  ppt_dat$ppt_grd %>%
    st_intersection(ngp, .) %>%
    filter(st_geometry_type(., by_geometry = TRUE) %in% c('MULTIPOLYGON', 'POLYGON'))  %>%
    mutate(anomaly = case_when(
      L3 == "Northwestern Glaciated Plains" ~ ppt - 447,  
      L3 == "Northwestern Great Plains" ~ ppt - 431 )) %>%
    ggplot() + theme_void() +
    geom_sf(aes(fill = anomaly), color = 'NA') +
    geom_sf_pattern(data =  ppt_dat$anom_map %>% filter(anom_cat == 'beyond'), 
                    aes(pattern = anom_cat), 
                    pattern_color = NA,
                    pattern_fill = "grey80",
                    pattern_angle = 45,
                    pattern_alpha = 0.5,
                    pattern_density = 0.5,
                    pattern_spacing = 0.025,
                    pattern_key_scale_factor = 1,, 
                    show.legend = F) +
    geom_sf(data = ngp_counties %>% 
  st_intersection(states %>% st_transform(st_crs(ngp_counties))), 
            fill = NA, color = 'grey70') + 
    geom_sf(data = states, fill = NA, color = 'black') + 
    geom_sf_text(data = states %>%
                   st_centroid(), 
                 aes(label = state),
                 fontface = 'bold',
                  color = 'black') + 
    geom_sf_label(data = locations %>% filter(type != 'Wildfire'), 
            aes(color = zone),
            label = expression(R[X]), 
            label.padding = unit(0.1, "lines"),
            label.r = unit(0.5, "lines"),
            label.size = 0.15,
            size = 5) +
    geom_sf(data = locations %>% filter(type == 'Wildfire'), 
            aes(shape = zone, color = zone), 
             stroke = 2, size = 3) +
    # geom_sf(data = locations, aes(shape = type, color = type), 
    #         fill = 'lightgreen', 
    #         stroke = 2, size = 3) +
    scale_fill_gradient2("Precipitation\nanomaly (mm)", 
                         low = wes_palette("Zissou1")[4],
                         mid = 'white',
                         high = wes_palette("Zissou1")[1], 
                         na.value = "grey70") +
    scale_pattern_manual(values = c(beyond = "circle", within = "none")) + 
    scale_shape_manual('Zone', values = c(4,3)) + 
    scale_color_manual('Fire type', values = c('darkgreen', 'darkred'), guide = 'none')

us_gg <- 
  ggplot() + theme_void() + 
 # geom_sf(data = usa %>% st_buffer(10000), color = NA, fill = 'white')  + 
  geom_sf(data = usa, color = 'white', fill = 'grey60')  + 
  geom_sf(data = gp, fill = 'white', 
          color =  'black', 
          alpha = 0.5, size = 2) +
  geom_sf(data = ngp, 
          fill =  wes_palette('Zissou1')[5],  
          color = NA, 
          show.legend = F) +
  scale_fill_manual('EPA Level III\necoregion', 
                    values = c("#74A089","#E1BD6D"))

grid.newpage() 
main <- viewport(width = 1, height = 0.95, x = 0.5, y = 0.5) 
inset <- viewport(width = 0.3, height = 0.4, x = 0.725, y = 0.7)
print(reg_gg, vp = main)
print(us_gg, vp = inset)  
```

```{r region_map}
reg_gg <-
ppt_dat$ngp_ppt %>%
  mutate(NA_L3NAME = str_replace(NA_L3NAME, ' ', '\n')) %>%
ggplot() + theme_void() +
  geom_sf(aes(fill = NA_L3NAME, alpha = ppt), 
          color = 'NA') +
  geom_sf(data = NGP, 
          fill = NA) + 
   geom_sf_label(data = locations %>% 
                  filter(type != 'Wildfire'), 
                label = expression(R[X]), 
                label.padding = unit(0.1, "lines"),
                label.r = unit(0.5, "lines"),
                label.size = 0.15,
                size = 4) +
  geom_sf(data = NGP_fires, 
          pch = 1, alpha = 0.5) + 
  geom_sf_label(data = NGP %>%
                  group_by(STATE_NAME) %>%
                  summarize(area = sum(Shape_Area)) , 
                aes(label = state.abb[match(STATE_NAME,state.name)]), 
                fontface='bold') + 
  scale_fill_manual('EPA Level III\necoregion', 
                    values = c("#74A089","#E1BD6D")) +
  scale_alpha_continuous("Precipitation\nmean annual (mm)", 
                         range = c(0.1, 1)) + 
  theme(legend.position = 'right', 
        legend.key.height = unit(1, 'cm')) + 
  guides(alpha = guide_legend(order = 2), fill = guide_legend(order = 1) )

us_gg <- 
  ggplot() + theme_void() + 
  geom_sf(data = usa %>% st_buffer(10000), color = NA, fill = 'white')  + 
  geom_sf(data = usa, color = 'white', fill = 'grey60')  + 
  geom_sf(data = gp, fill = 'white', 
          color =  'black', 
          alpha = 0.5, size = 2) +
  geom_sf(data = NGP, 
          fill =  wes_palette('Zissou1')[5],  
          color = NA, 
          show.legend = F) +
  scale_fill_manual('EPA Level III\necoregion', 
                    values = c("#74A089","#E1BD6D"))

grid.newpage() 
main <- viewport(width = 1, height = 0.95, x = 0.5, y = 0.5) 
inset <- viewport(width = 0.3, height = 0.4, x = 0.25, y = 0.25)
print(reg_gg, vp = main)
print(us_gg, vp = inset)
```

```{r}
NGP_fires %>% 
  as_tibble() %>%
  group_by(STATE_NAME, NA_L3NAME) %>% 
  summarize(Fires = n()) %>%
  pivot_wider(names_from = NA_L3NAME, 
              values_from = Fires) %>%
   mutate(Total = `Northwestern Great Plains` + `Northwestern Glaciated Plains`) %>%
  #xtable::xtable() 
  pander::pander("Fire incidents within each LIII ecoregion by state.")
```

@@

```{r}
BurnSeverityData %>%
  group_by(zone, type) %>%
  summarize(fires = n() ) %>%
  pivot_wider(names_from = zone, 
              values_from = fires) %>%
  mutate(Total = east + west) %>%
  pander::pander("Incidents used in this study.")
```

## Burn severity across zones and fire types

```{r WF_v_Rx, fig.width = 8, fig.height = 4, fig.cap = " "}
    BurnSeverityData  %>% 
      mutate(zone = fct_rev(zone), 
             zone = paste0(str_to_title(zone), 'ern zone'),
             type = fct_relevel(type, c("Wildfire",'Rx' )), 
             type = recode(type, 'Rx'= "Prescribed\nburn")) %>%
      ggplot(aes(x= type, y = dNBR_Mean)) + theme_bw(16) +
       geom_boxplot(varwidth = TRUE, 
                    fill = 'lightblue', 
                    staplewidth = 0.25) +
       geom_point(data = . %>%
                   group_by(type, zone) %>%
                   summarize(Mean = mean(dNBR_Mean)), 
                 aes(y = Mean), 
                 pch = 24, size = 4, 
                 color = 'white', fill = 'darkblue') + 
      labs(x = 'Fire type', 
           y = 'Burn severity (ΔNBR)') +           
      facet_wrap(~zone, scales = 'free_x') +
      theme(panel.grid.major.x = element_blank(), 
            axis.text.x = element_text(color = 'black'))
```

>>>>

---
title: 'Remote sensing in rangeland fire ecology: Comparing imagery to measured fire behavior, and burn severity across prescribed burns and wildfires'
subtitle:  'Data and analyses'
author: "DAM"
date: "Last compiled: `r Sys.Date()`"
output:
  html_document:
    theme: united
    highlight: tango
    toc: true
    toc_float: true
---

```{r setup, echo=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning= FALSE,
                      message= FALSE, eval = TRUE)
                      #fig.path = "./, dev = c('png', 'pdf')) 
pacman::p_load(tidyverse, sf, ggpattern, pander, readxl, grid, gridExtra)
pacman::p_load_gh('devanmcg/wesanderson')
```

@@

```{r region_map}
ggplot() + theme_void() +
  geom_sf(data = NGP, 
          aes(fill = NA_L3NAME), 
          alpha = 0.6) +
  geom_sf(data = NGP_fires, 
          pch = 1, alpha = 0.5) + 
  geom_sf_label(data = NGP %>%
                 group_by(STATE_NAME) %>%
                 summarize(area = sum(Shape_Area)) , 
               aes(label = state.abb[match(STATE_NAME,state.name)])) + 
  scale_fill_manual('L3 ecoregion', 
                    values = c("#74A089","#E1BD6D")) +
  theme(legend.position = 'inside', 
        legend.position.inside = c(0.3, 0.23))
```

@@

  ppt_dat$ppt_grd %>%
    st_intersection(ngp, .) %>%
    filter(st_geometry_type(., by_geometry = TRUE) %in% c('MULTIPOLYGON', 'POLYGON'))  %>%
    mutate(anomaly = case_when(
      L3 == "Northwestern Glaciated Plains" ~ ppt - 447,  
      L3 == "Northwestern Great Plains" ~ ppt - 431 )) %>%
    ggplot() + theme_void() +
    geom_sf(aes(fill = anomaly), color = 'NA') +
    geom_sf_pattern(data =  ppt_dat$anom_map %>% filter(anom_cat == 'beyond'), 
                    aes(pattern = anom_cat), 
                    pattern_color = NA,
                    pattern_fill = "grey80",
                    pattern_angle = 45,
                    pattern_alpha = 0.5,
                    pattern_density = 0.5,
                    pattern_spacing = 0.025,
                    pattern_key_scale_factor = 1,, 
                    show.legend = F) +
    geom_sf(data = ngp_counties, fill = NA, color = 'grey70') + 
    geom_sf(data = states, fill = NA, color = 'black') + 
    geom_sf_text(data = states %>%
                   st_centroid(), 
                 aes(label = state),
                 fontface = 'bold',
                  color = 'black') + 
    geom_sf_label(data = locations %>% filter(type != 'Wildfire'), 
            aes(color = zone),
            label = expression(R[X]), 
            label.padding = unit(0.1, "lines"),
            label.r = unit(0.5, "lines"),
            label.size = 0.15,
            size = 5) +
    geom_sf(data = locations %>% filter(type == 'Wildfire'), 
            aes(shape = zone, color = zone), 
             stroke = 2, size = 3) +
    # geom_sf(data = locations, aes(shape = type, color = type), 
    #         fill = 'lightgreen', 
    #         stroke = 2, size = 3) +
    scale_fill_gradient2("Precipitation\nanomaly (mm)", 
                         low = wes_palette("Zissou1")[4],
                         mid = 'white',
                         high = wes_palette("Zissou1")[1], 
                         na.value = "grey70") +
    scale_pattern_manual(values = c(beyond = "circle", within = "none")) + 
    scale_shape_manual('Zone', values = c(4,3)) + 
    scale_color_manual('Fire type', values = c('darkgreen', 'darkred'), guide = 'none')
```

@@

```{r WF_v_Rx, fig.width = 8, fig.height = 4, fig.cap = " "}
    BurnSeverityData  %>% 
      mutate(zone = ifelse(L3 == "Northwestern Great Plains", 
                           'Western zone', 'Eastern zone'), 
             zone = fct_rev(zone), 
             type = fct_relevel(type, c("Wildfire",'Rx' )), 
             type = recode(type, 'Rx'= "Prescribed\nburn")) %>%
      ggplot(aes(x= type, y = dNBR_Mean)) + theme_bw(16) +
       geom_boxplot(varwidth = TRUE, 
                    fill = 'lightblue', 
                    staplewidth = 0.25) +
       geom_point(data = . %>%
                   group_by(type, zone) %>%
                   summarize(Mean = mean(dNBR_Mean)), 
                 aes(y = Mean), 
                 pch = 24, size = 4, 
                 color = 'white', fill = 'darkblue') + 
      labs(x = 'Fire type', 
           y = 'Burn severity (ΔNBR)') +           
      facet_wrap(~zone, scales = 'free_x') +
      theme(panel.grid.major.x = element_blank(), 
            axis.text.x = element_text(color = 'black'))
```
